\HeaderA{bdendo}{A case-control study of endometrial cancer}{bdendo}
\keyword{datasets}{bdendo}
\begin{Description}\relax
The \code{bdendo} data frame has 315 rows and 13 columns.
These data concern a study in which each case of endometrial cancer was
matched with 4 controls. Matching was by date of birth (within one
year), marital status, and residence.
\end{Description}
\begin{Format}\relax
This data frame contains the following columns:
\Tabular{rl}{
\code{set}: & Case-control set: a numeric vector  \\
\code{d}: & Case or control: a numeric vector (1=case, 0=control) \\
\code{gall}: & Gall bladder disease: a factor with levels
\code{No} 
\code{Yes}. \\
\code{hyp}: & Hypertension: a factor with levels
\code{No} 
\code{Yes}. \\
\code{ob}: & Obesity: a factor with levels
\code{No} 
\code{Yes}. \\
\code{est}: & A factor with levels
\code{No} 
\code{Yes}. \\
\code{dur}: & Duration of conjugated oestrogen therapy: an ordered factor with levels
\code{0} < \code{1} < \code{2} < \code{3} < \code{4}. \\
\code{non}: & Use of non oestrogen drugs: a factor with levels
\code{No} 
\code{Yes}. \\
\code{duration}: & Months of oestrogen therapy: a numeric vector.  \\
\code{age}: & A numeric vector. \\
\code{cest}: & Conjugated oestrogen dose: an ordered factor with levels
\code{0} < \code{1} < \code{2} < \code{3}. \\
\code{agegrp}: & A factor with levels
\code{55-59} 
\code{60-64} 
\code{65-69} 
\code{70-74} 
\code{75-79} 
\code{80-84} \\
\code{age3}: & a factor with levels
\code{<64} 
\code{65-74} 
\code{75+}  \\
}
\end{Format}
\begin{Source}\relax
Breslow NE, and Day N, Statistical Methods in Cancer Research. Volume
I: The Analysis of Case-Control Studies. IARC Scientific
Publications, IARC:Lyon, 1980.
\end{Source}
\begin{Examples}
\begin{ExampleCode}
data(bdendo)
\end{ExampleCode}
\end{Examples}

