\HeaderA{expand.data}{Function to expand data for regression analysis of interval censored
data.}{expand.data}
\keyword{models}{expand.data}
\keyword{regression}{expand.data}
\keyword{survival}{expand.data}
\begin{Description}\relax
This is a utility function.

The original records with \code{first.well}, \code{last.well} and
\code{first.ill} are
expanded to multiple records; several for each interval where the
person is known to be well and one where the person is known to fail.
At the same time columns for the covariates needed to estimate rates
and the response variable are generated.
\end{Description}
\begin{Usage}
\begin{verbatim}
expand.data(fu, formula, breaks, data)
\end{verbatim}
\end{Usage}
\begin{Arguments}
\begin{ldescription}
\item[\code{fu}] A 3-column matrix with \code{first.well}, \code{last.well} and
\code{first.ill} in each row.
\item[\code{formula}] Model fromula, used to derive the model matrix.
\item[\code{breaks}] Defines the intervals in which the baseline rate is
assumed constant. All follow-up before the first and after the
last break is discarded.
\item[\code{data}] Datafrem in which \code{fu} and \code{formula} is interpreted.
\end{ldescription}
\end{Arguments}
\begin{Value}
Returns a list with three components
\begin{ldescription}
\item[\code{rates.frame}] Dataframe of covariates for estimation of the
baseline rates --- one per interval defined by \code{breaks}.
\item[\code{cov.frame}] Dataframe for estimation of the covariate effects. A
data-framed version of the designmatrix from \code{formula}.
\item[\code{y}] Response vector.
\end{ldescription}
\end{Value}
\begin{Author}\relax
Martyn Plummer, \email{plummer@iarc.fr}
\end{Author}
\begin{References}\relax
B Carstensen: Regression models for interval censored
survival data: application to HIV infection in Danish homosexual
men. Statistics in Medicine, 15(20):2177-2189, 1996.
\end{References}
\begin{SeeAlso}\relax
\code{\LinkA{Icens}{Icens}}
\code{\LinkA{fit.mult}{fit.mult}}
\code{\LinkA{fit.add}{fit.add}}
\end{SeeAlso}
\begin{Examples}
\begin{ExampleCode}
  \end{ExampleCode}
\end{Examples}

