\HeaderA{gmortDK}{Population mortality rates for Denmark in 5-years age groups.}{gmortDK}
\keyword{datasets}{gmortDK}
\begin{Description}\relax
The \code{gmortDK} data frame has 418 rows and 21 columns.
\end{Description}
\begin{Format}\relax
This data frame contains the following columns:
\Tabular{rl}{
\code{agr}: & Age group, 0:0--4, 5:5--9,..., 90:90+. \\
\code{per}: & Calendar period, 38: 1938--42, 43: 1943--47, ..., 88:1988-92. \\
\code{sex}: & Sex, 1: male, 2: female. \\
\code{risk}: & Number of person-years in the Danish population. \\
\code{dt}: & Number of deaths. \\
\code{rt}: & Overall mortality rate in cases per 1000 person-years, i.e.
\code{rt=1000*dt/risk} \\
& Cause-specific mortality rates in cases per 1000 person-years: \\
\code{r1}: & Infections \\
\code{r2}: & Cancer. \\
\code{r3}: & Tumors, benign, unspecific nature. \\
\code{r4}: & Endocrine, metabolic. \\
\code{r5}: & Blood. \\
\code{r6}: & Nervous system, psychiatric. \\
\code{r7}: & Cerebrovascular. \\
\code{r8}: & Cardiac. \\
\code{r9}: & Respiratory diseases, excl. cancer. \\
\code{r10}: & Liver, excl. cancer. \\
\code{r11}: & Digestive, other. \\
\code{r12}: & Genitourinary. \\
\code{r13}: & Ill-defined symptoms. \\
\code{r14}: & All other, natural. \\
\code{r15}: & Violent. \\
}
\end{Format}
\begin{Source}\relax
Statistics Denmark, National board of health provided original data. 
Michael Andersson grouped the causes of death.
\end{Source}
\begin{SeeAlso}\relax
\code{\LinkA{thoro}{thoro}}, \code{\LinkA{mortDK}{mortDK}}
\end{SeeAlso}
\begin{Examples}
\begin{ExampleCode}
data(gmortDK)
\end{ExampleCode}
\end{Examples}

