\HeaderA{S.typh}{Salmonella Typhimurium outbreak 1996 in Denmark.}{S.typh}
\keyword{datasets}{S.typh}
\begin{Description}\relax
Matched case-control study of food poisoning.
\end{Description}
\begin{Format}\relax
A data frame with 136 observations on the following 15 variables:
\Tabular{rl}{
\code{id}: & Person identification \\
\code{set}: & Matched set indicator \\
\code{case}: & Case-control status (1:case, 0:control \\
\code{age}: & Age of individual \\
\code{sex}: & Sex of individual (1:male, 2:female) \\ 
\code{abroad}: & Within the last two weeks visited abroad (1:yes, 0:no) \\
\code{beef}: & Within the last two weeks eaten beef \\
\code{pork}: & Within the last two weeks eaten pork \\
\code{veal}: & Within the last two weeks eaten veal \\
\code{poultry}: & Within the last two weeks eaten poultry \\
\code{liverp}: & Within the last two weeks eaten liverpaste \\
\code{veg}: & Within the last two weeks eaten vegetables \\
\code{fruit}: & Within the last two weeks eaten fruit \\
\code{egg}: & Within the last two weeks eaten eggs \\
\code{plant7}: & Within the last two weeks eaten meat from plant no. 7 \\
}
\end{Format}
\begin{Details}\relax
In the fall of 1996 an unusually large number of Salmonella
Typhimurium cases were recorded in Fyn county in Denmark. The Danish
Zoonosis Centre set up a matched case-control study to find the
sources. Cases and two age-, sex- and residency-matched controls were
telephone interviewed about their food intake during the last two
weeks.

The participants were asked at which retailer(s) they had purchased
meat. Retailers were independently of this linked to meat processing
plants, and thus participants were linked to meat processing
plants. This way persons could be linked to (amongst other) plant no 7.
\end{Details}
\begin{Source}\relax
Tine Hald.
\end{Source}
\begin{References}\relax
Mølbak K and Hald T: Salmonella Typhimurium outbreak in late summer
1996. A Case-control study. (In Danish:
Salmonella typhimurium udbrud på Fyn sensommeren
1996. En case-kontrol undersøgelse.) Ugeskrift for Læger.,
159(36):5372-7, 1997.
\end{References}
\begin{Examples}
\begin{ExampleCode}
data(S.typh)
\end{ExampleCode}
\end{Examples}

