\HeaderA{thoro}{Thorotrast Study}{thoro}
\keyword{datasets}{thoro}
\begin{Description}\relax
The \code{thoro} data frame has 2470 rows and 14 columns.
Each row represents one patient that have had cerebral angiography (X-ray of
the brain) with an injected contrast medium, either Thorotrast or
another one (the controls).
\end{Description}
\begin{Format}\relax
This data frame contains the following columns:
\Tabular{rl}{
\code{id}:       & Identification of person. \\
\code{sex}:      & Sex, 1: male / 2: female. \\
\code{birthdat}: & Date of birth, \code{Date} variable. \\
\code{contrast}: & Group, 1: Thorotrast / 2: Control. \\
\code{injecdat}: & Date of contrast injection, \code{Date} variable. \\
\code{volume}:   & Injected volume of Thorotrast in ml. Control
patients have a 0 in this variable. \\
\code{exitdat}:  & Date of exit from the study, \code{Date} variable. \\
\code{exitstat}: & Status at exit, 1: dead / 2: alive,
censored at closing of study, 20 February 1992 / 3:
censored alive at some earlier date. \\
\code{cause}:    & Cause of death. See causes in the helpfile for
\code{\LinkA{gmortDK}{gmortDK}} \\
\code{liverdat}: & Date of liver cancer diagnosis, \code{Date} variable. \\
\code{liver}:    & Indicator of liver cancer diagnosis. Not all livercancers
are histologically verified, hence
\code{liver >= hepcc + chola + hmang} \\
\code{hepcc}:    & Hepatocellular carcinoma at \code{liverdat}. \\
\code{chola}:    & Cholangiocellular carcinoma at \code{liverdat}. \\
\code{hmang}:    & Haemangisarcoma carcinoma at \code{liverdat}. \\
}
\end{Format}
\begin{Source}\relax
M Andersson, M Vyberg, J Visfeldt, B Carstensen \& HH Storm:
Primary liver tumours among Danish patients exposed to Thorotrast.
Radiation Research, 137, pp. 262--273, 1994.

M Andersson, B Carstensen HH Storm: Mortality and cancer
incidence after cerebral angiography. Radiation Research, 142,
pp. 305--320, 1995.
\end{Source}
\begin{SeeAlso}\relax
\code{\LinkA{mortDK}{mortDK}}, \code{\LinkA{gmortDK}{gmortDK}}
\end{SeeAlso}
\begin{Examples}
\begin{ExampleCode}
data(thoro)
str(thoro)
\end{ExampleCode}
\end{Examples}

